### 2021-05-31 paper/parlr_paper_1.Rmd
wss slides first draft

## Introduction (or Abstract?)

Record linkage is the task of identifying duplicate records across multiple data sources. This task is straightforward when provided unique identifiers or highly reliable information, but becomes difficult when data sources are plagued with error. Additionally, record linkage is inherently computationally intensive, and many existing methods are not feasible for many applications. In this paper, we propose a Bayesian model for record linkage that is amenable to parallel computing, hashing techniques and efficient storage mechanisms, that is scalable to large linkage tasks.

@@

Most record linkage techniques are derived from the seminal 1969 paper by Fellegi and Sunter: "A Theory for Record Linkage". The defining characteristic of their model was to transform the sets of records, which often contain text data that is difficult to model, into sets of comparison vectors governed by parameters that can be more easily estimated. Concretely, if files $A$ and $B$ have $n_A$ and $n_B$ records respectively, and if the files share $F$ fields in common upon which to base the linkage, the Fellegi and Sunter approach generates a $n_A n_B \times F$ matrix $\Gamma$, which contains similarity scores between each pair of records across datasets. We say $\gamma_{ij}$ is the comparison vector for record $i \in A$ and record $j \in B$, with $\gamma_{ij}^f$ providing their similarity score on the $f^{th}$ field. For ease of modeling and computation, we restrict these similarity scores to be discrete, ordinal variables, and the construction of these is left to the modeler. It is common to use use binary 0-1 variables to indicate exact matching, and 0-1-2 variables to provide an option for partial matching. For text data, we calculate similarity based on Levenstein distance or some other text similarity score, and bin these scores to integers for use in the model.

Two modern adaptations of the Fellegi Sunter model are important for understanding `parlr` and its contribution. In 2019, Enamorado et al proposed `fastlink`, a method that closely followed the modelling assumptions of Fellegi and Sunter, but used innovating hashing techniques to greatly enhance its speed and scale. Importantly, they maintained Fellegi and Sunter's independent matching assumption, in which the matching status of any $(i,j)$ pair is made independently of the matching of any other $(i', j')$. This leads to many matchings that violate "one-to-one" considerations that are often implicit in the data, which need to be resolved in a post-processing step. In contrast, Sadinle (2017) proposed "Beta Record Linkage for Bipartite Matching" and the accompanying `BRL` package which strictly enforces one-to-one matching. Specifically, in each iteration of his Gibbs sampler, he considers each record $j\in B$, removes from consideration the records $i\in A$ that have already been matched, then samples a potential link. However, accounting for these dependencies throughout the linkage process is computationally burdensome, leaving `BRL` only suitable for small to moderate linkage problems.

Outside of the Fellegi-Sunter framework, some researchers have developed methods that model the data directly, not a derived set of comparison vectors. As an early example, Steorts et al 2016 produced `blink`, which had the advantage of a computational complexity that grew linearly with the data (as opposed to quadratically within Fellegi Sunter), but was plagued with slow mixing times and was unable to incorporate text data. Later, Steorts et al 2020 improved on this method with `dblink`, which was able to simulate text data within its Gibbs sampler by drawing from an empirical distribution, and remedied its slow mixing through a probabilistic blocking approach. However, this method remains computationally intensive and requires distributed computing to manage even moderate linkage problems in acceptable time.

@@

Our notation and assumptions closely follow that of Sadinle (2017). Denote two files as $A$ and $B$, with $n_A$ and $n_B$ records respectively, and with records indexed as $i \in \{1, \ldots, n_A\}$ in $A$ and $j \in \{1, \ldots, n_B\}$ in $B$. Without loss of generality, label the files such that $n_A \geq n_B$. We also assume there are no duplicates within files, only across. For each record pair under consideration, we generate a comparison vector $\boldsymbol{\gamma}_{ij} = \{\gamma_{ij}^1, \ldots, \gamma_{ij}^F\}$, where  $F$ is the number of fields used in the linkage and each $\gamma_{ij}^f$ takes on a value $l \in \{1, \ldots, L_f\}$ indicating the level agreement between the two records on a specified field.

@@

# Model Specification

@@

Here the reader should the relationship between our proposed model and that of Sadinle 2017. In his model, Sadinle constructs a prior distribution on the entire $\mathbf{Z}$ vector, which induces a Gibbs sampler that strictly enforces one-to-one matching. In particular, this sampler removes previously matches records from the set of candidate records when sampling $Z_j$, creating a dependency that makes the sampler *inherently serial*. We however use independent priors for each $Z_j$, thereby weakening the one-to-one requirement; our sampler does ensure that each record in $B$ can be matched to at most one record in $A$, but allows for the possiblity that multiple records in $B$ match to the same record in $A$. With this change, components of the $\mathbf{Z}$ vector can be computed in parallel, allowing for significant computational gains.  More importantly, since only the agreement pattern of $Z_j$ is used for calculations within the Gibbs sampler, and not the particular record label, we can conduct this sampling only at the level of the unique agreement patterns, offering even more computational savings. This will be explained further in \textcolor{red}{Section XX}.

@@

@@

To address this, we propose a new strategy called *storage efficient indexing* (SEI). In standard indexing, the modeler decides a certain criteria that they expect all true matching pairs to satisfy, and a priori label any record pairs that do not meet that criteria as non-matching. For example, one might only consider pairs with a certain similarity score on a field deemed to be important (like first name), or only pairs with exact matching on a specified number of fields. While generally chosen to be be quite loose, establishing these criteria requires knowledge of the problem and invites room for human error. We propose a method of reducing the comparison space (and reducing storage requirements) without these drawbacks. Note that all records of the same agreement pattern have the same probability when sampling $Z_j$. Therefore we know that records belonging to an $h_p$ such that $H_{p_j}$ is large are very unlikely to be sampled consistently enough to be deemed a match through the Bayes estimate, even without considering the form of the agreement pattern itself. In random indexing, rather than store all of these unlikely record labels, we choose to store only a small number $R$ of them. Posterior calculations still attribute the appropriate weight to all records through the summary statistics $H_{p_j}, H_p^m$, and $H_p^u$; this is similar to the *u-correction* concept introduced by ________. Rather than storing $n_A \times n_B$ record labels, random indexing allows us to store at most $n_B \times P \times R$, regardless of how large $n_A$ is. Coupled with the partitioning strategy for creating comparison vectors, this greatly expands the size of the linkage tasks we can undertake.

@@

# Simulation Studies

In his 2017 paper, Sadinle demonstrated the strength of his method by running `BRL` on simulated pairs of files with differing levels of errors and overlap. Comparing our method against `BRL` on these same simulated datasets, we find that our method only has weakened performance in the extreme scenario of very high errors and very high overlap across files. We use the following measures of accuracy in our analysis: although using only recall and precision is common in the literature, we find it useful to have a single metric the gauge overall performance, and thus also include F measure.

@@

To demonstrate speed, we generate comparison vectors from pre-specified distributions so that we can easily increase the size of the linkage problem. Distributions are meant to emulate the behavior of similarity scores across first name, last name, and day, month, and and year of birth, with 1 indicating an exact match on a particular field and 0 indicating a nonmatch. We simulate these data for different values of $n_A$ and $n_B$, and compare the run-time of `parlr` against `BRL`. We see that at low data size, `BRL` outperforms, but that `parlr` is significantly faster at handling larger data.

In particular, run-time for `BRL` seems to grow quadratically while run-time for `parlr` seems to grow linearly.

@@

<!-- ```{r fig.align='center'} -->
<!-- knitr::include_graphics("../notes/figures/speed_distributions.png") -->
<!-- ``` -->

@@

The above discussion suggests that for fixed $n_B$, computation time should remain mostly constant with growing $n_A$. Shockingly, this seems to be true. In the plot below, fixing $n_B = 500$, we see linear growth for the run-time under `BRL` as $n_A$ increases, with much more static run-time under `parlr`. The slight increases in run-time that we do see are due primarily to the hashing step, which again can be run in parallel for large data.

@@

- Thinking of re-running the El Salvador linkage from the Sadinle 2017 paper, but doing so with without the formal blocking and indexing that he uses. Instead, I'd use hashing and storage efficient indexing -- I should be able to do linkage on the all the data at once!

@@

<!-- We note that this often not a reasonable assumption; for example if we know that two records differ on gender, this does not give any information on whether their year of births are the same, but it does suggest that its more likely that their first names are different.  -->

>>>>

## Introduction

Record linkage is the task of identifying duplicate records across multiple data sources. This task is straightforward when provided unique identifiers or highly reliable information, but becomes difficult when data sources are plagued with error. Additionally, many record linkage tasks are large in nature, and many existing methods are too computational intensive to be used in practice. In this paper, we propose a Bayesian model for record linkage that is amenable to parallel computing, hashing techniques and efficient storage mechanisms, that is scalable to large linkage tasks.

The proposed method is inspired by the "Beta Record Linkage" method proposed by Mauricio Sadinle's 2017 "Bayesian Estimation of Bipartite Matching for Record Linkage". As such, we demonstrate our method through a comparison of our methods against Sadinle's on the simulation study he designed and the El Salvador homocide case study he utilized. We include one additional simulation study to demonstrate the power of our method on large datasets. In what follows, Section 2 describes previous work in record linkage that lays the foundation for our method, Section 3 provides our model and the Gibbs sampler used for posterior inference, and Section 4 describes strategies used for more efficient and scalable implementation. Then, Sections 5 and 6 demonstrate the method through simulations and a case study. Lastly, our discussion highlights open questions in the record linkage liturature and motivates future work.

@@

Most record linkage techniques are derived from the seminal 1969 paper by Fellegi and Sunter: "A Theory for Record Linkage". The defining characteristic of their model was to transform the sets of records, which often contain text data that is difficult to model, into sets of comparison vectors governed by parameters that can be more easily estimated. Concretely, if files $A$ and $B$ have $n_A$ and $n_B$ records respectively, and if the files share $F$ fields in common upon which to base the linkage, the Fellegi and Sunter approach generates a $n_A n_B \times F$ matrix $\Gamma$, which contains similarity scores between each pair of records across datasets. We say $\gamma_{ij}$ is the comparison vector for record $i \in A$ and record $j \in B$, with $\gamma_{ij}^f \in \{1, \ldots, L_f\}$ providing their similarity score on the $f^{th}$ field. For ease of modeling and computation, we restrict these similarity scores to be discrete, ordinal variables, and the construction of these is left to the modeler. We adopt the convention that $\gamma_{ij}^f = 1$ corresponds to the highest level of agreement and $\gamma_{ij}^f = L_f$ corresponds to the lowest. It is common to use use binary 1-2 variables to indicate exact matching, and 1-2-3 variables to provide an option for partial matching. For text data, we calculate similarity based on Levenstein distance or some other text similarity score, and bin these scores to integers for use in the model.

The likelihood in the original Fellegi Sunter model is a mixture model through which each record pair is independently classified as a match or nonmatch. This independent classification often leads to sets of matches that break expectations of transisitivity, and Jaro devised an procedure to reduce a set of conflicting matches to the mostly likely set of one-to-one matchings. One key strength of Sadinle's `BRL` is that it is garunteed to produce a set of bipartite matchings that respect one-to-one assumptions--without the need for any post-hoc procedures. Specifically, in each iteration of his Gibbs sampler, he considers each record $j\in B$, removes from consideration the records $i\in A$ that have already been matched, and then samples a potential link. However, accounting for these dependencies throughout the linkage process is computationally burdensome, leaving `BRL` only suitable for small to moderate linkage problems.

<!-- Two modern adaptations of the Fellegi Sunter model are important for understanding `parlr` and its contribution. In 2019, Enamorado et al proposed `fastlink`, a method that closely followed the modelling assumptions of Fellegi and Sunter, but used innovating hashing techniques to greatly enhance its speed and scale. Importantly, they maintained Fellegi and Sunter's independent matching assumption, in which the matching status of any $(i,j)$ pair is made independently of the matching of any other $(i', j')$. This leads to many matchings that violate "one-to-one" considerations that are often implicit in the data, which need to be resolved in a post-processing step. In contrast, Sadinle (2017) proposed "Beta Record Linkage for Bipartite Matching" and the accompanying `BRL` package which strictly enforces one-to-one matching. Specifically, in each iteration of his Gibbs sampler, he considers each record $j\in B$, removes from consideration the records $i\in A$ that have already been matched, then samples a potential link. However, accounting for these dependencies throughout the linkage process is computationally burdensome, leaving `BRL` only suitable for small to moderate linkage problems.  -->

Outside of the Fellegi-Sunter framework, some researchers have developed methods that model the data directly, not a derived set of comparison vectors. As an early example, Steorts et al 2016 produced `blink`, which had the advantage of a computational complexity that grew linearly with the data (as opposed to quadratically within Fellegi Sunter), but was plagued with slow mixing times and was unable to incorporate text data. Later, Steorts et al 2020 improved on this method with `dblink`, which was able to simulate text data within its Gibbs sampler by drawing from an empirical distribution, and remedied its slow mixing through a probabilistic blocking approach. However, this method remains computationally intensive and requires distributed computing to manage even moderate linkage problems in acceptable time. Because these methods fall outside of the Fellegi-Sunter framework, we do not consider them further in this paper.

@@

Our notation and assumptions closely follow those of Sadinle. Denote two files as $A$ and $B$, with $n_A$ and $n_B$ records respectively, and with records indexed as $i \in \{1, \ldots, n_A\}$ in $A$ and $j \in \{1, \ldots, n_B\}$ in $B$. Without loss of generality, label the files such that $n_A \geq n_B$. We also assume there are no duplicates within files, only across. For each record pair under consideration, we generate a comparison vector $\boldsymbol{\gamma}_{ij} = \{\gamma_{ij}^1, \ldots, \gamma_{ij}^F\}$, where  $F$ is the number of fields used in the linkage and each $\gamma_{ij}^f$ takes on a value $l \in \{1, \ldots, L_f\}$ indicating the level agreement between the two records on a specified field.

@@

## Model Specification

@@

Here the reader should the relationship between our proposed model and that of Sadinle 2017. In his model, Sadinle constructs a prior distribution on the entire $\mathbf{Z}$ vector, which induces a Gibbs sampler that strictly enforces one-to-one matching. In particular, this sampler removes previously matches records from the set of candidate records when sampling $Z_j$, creating a dependency that makes the sampler *inherently serial*. We however use independent priors for each $Z_j$, thereby weakening the one-to-one requirement; our sampler does ensure that each record in $B$ can be matched to at most one record in $A$, but allows for the possiblity that multiple records in $B$ match to the same record in $A$. We explore the ramifications of this distinction with the El Salvador case study. With this change, components of the $\mathbf{Z}$ vector can be computed in parallel, allowing for significant computational gains.  More importantly, since only the agreement pattern of $Z_j$ is used for calculations within the Gibbs sampler, and not the particular record label, we can conduct this sampling only at the level of the unique agreement patterns, offering even more computational savings. This will be explained further in \textcolor{red}{Section XX}.

@@

We calculate a Bayes estimate $\hat{\mathbf{Z}}$ for the linkage parameter $\mathbf{Z}$ by assigning different positive losses to different types of errors, and minimizing posterior loss. Specifically, we minimize the quantity $L(\hat{\mathbf{Z}}, \mathbf{Z}) = \sum_{j=1}^{n_B} L(\hat{Z_j}, Z_j)$ where

$$L(\hat{Z_j}, Z_j)=\begin{cases} 
    0  & \text{if } Z_j = \hat{Z_j}; \\
    n_A + 1,  & \text{if record } j\in B \text{ does not have a match in file } A; \\
\end{cases}$$

@@

To address this, we propose a new strategy called *storage efficient indexing* (SEI). In standard indexing, the modeler decides a certain criteria that they expect all true matching pairs to satisfy, and a priori label any record pairs that do not meet that criteria as non-matching. For example, one might only consider pairs with a certain similarity score on a field deemed to be important (like first name), or only pairs with exact matching on a specified number of fields. While generally chosen to be be quite loose, establishing these criteria requires knowledge of the problem and invites room for human error. We propose a method of reducing the comparison space (and reducing storage requirements) without these drawbacks. Note that all records of the same agreement pattern have the same probability when sampling $Z_j$. Therefore we know that records belonging to an $h_p$ such that $H_{p_j}$ is large are very unlikely to be sampled consistently enough to be deemed a match through the Bayes estimate, even without considering the form of the agreement pattern itself. In random indexing, rather than store all of these unlikely record labels, we choose to store only a small number $R$ of them. Posterior calculations still attribute the appropriate weight to all records through the summary statistics $H_{p_j}, H_p^m$, and $H_p^u$. Rather than storing $n_A \times n_B$ record labels, SEI allows us to store at most $n_B \times P \times R$ labels, regardless of how large $n_A$ is. Coupled with the partitioning strategy for creating comparison vectors, this greatly expands the size of the linkage tasks we can undertake.

@@

# Simulation Study 1 - Accuracy

In his 2017 paper, Sadinle demonstrated the accuracy of his method by running `BRL` on simulated pairs of files with differing levels of errors and overlap. Each file consists of first name, last name, gender, postal code, age, and occupation, and following Sadinle's implementation, we use only four fields for the linkage process: both names, age, and occupation. Similarity scores for the names are binned into three levels according the default settings for the `compareRecords` function in the `BRL` package, and age and occupation similarity scores are binary for exact matching. We use flat priors for all $\mathbf{m}$ and $\mathbf{u}$ parameters, and run the Gibbs sampler for 1000 iterations, with the first 100 discarded as burn-in.

Comparing our method against `BRL` on these same simulated datasets, we find that our method only has weakened performance in the most extreme scenario of very high errors and very high overlap across files. We use the following measures of accuracy in our analysis: although using only recall and precision is common in the literature, we find it useful to have a single metric the gauge overall performance, and thus also include F measure.

@@

# Simulation Study 2 - Speed

To demonstrate speed, we generate comparison vectors from pre-specified distributions so that we can easily increase the size of the linkage problem. Distributions are meant to emulate the behavior of similarity scores across first name, last name, and day, month, and and year of birth, with 1 indicating an exact match on a particular field and 0 indicating a nonmatch. We simulate these data for different values of $n_A$ and $n_B$, and compare the run-time of `parlr` against `BRL`. Note that the number of unique patterns $P$ is bounded above by $2^5 = 32$, a bound which is consistently attained in the larger simulations.

We see that at low data size, `BRL` outperforms, but that `parlr` is significantly faster at handling larger data. In particular, run-time for `BRL` seems to grow quadratically (or linearly with the size of both $A$ and $B$) while run-time for `parlr` seems to grow linearly (in the size of $B$).

@@

@@

The above discussion suggests that for fixed $n_B$, computation time should remain mostly constant with growing $n_A$. Simulation study suggests that this is true. In the plot below, fixing $n_B = 500$, we see linear growth for the run-time under `BRL` as $n_A$ increases, with much more static run-time under `parlr`. The slight increases in run-time that we do see are due primarily to the hashing step, which again can be run in parallel for large data.

@@

The country of El Salvador was immersed in civil war from 1980 to 1991, and throughout the time, several organizations attempted to document casualties of the conflict. When estimating the total number of casualities, one cannot simply sum the numbers recorded by each organization, as it is likely that the same individuals are recorded in multiple casuality lists. To obtain a more accurate estimate of the casualties then, we follow Sadinle's 2017 paper and link records of El Salvadoran civilian casualties from two sources: El Rescate - Tutela Regal (ERTL) and the Salvadoran Human Rights Commission (CDHES, by its acronym in Spanish). The ERTL dataset consists of digitized reports that had been published throughout the conflict. The CDHES dataset consists of casualties that had been reported directly to the organization, and later digitized.

There are several challenges with working with such data. Firstly, both datasets have been automatically digitized, which inherently leads to some degree of typographical error. Secondly, the CDHES records are all second hand accounts reported by individuals, which can result in additional errors. Lastly, the only fields recorded are given name, last name, date of death, and place of death; it is relatively common for a parent and child to share the same given name, resulting in indistinguishable records for two different individuals. This last point nearly breaks the above mentioned assumption that there are no duplicates within files, and reveals a key difference between `BRL` and our proposed method.

We only utilized records with nonmissing entries for given and last name, results in $n_A = 4420$ files in CHDES and $n_B = 1323$ files in ERTL. The names were standardized to account for common misspellings in the Spanish language, and then compared using a modified Levenstein distance to account for the fact that second names are often ommited. Place of birth is recorded by municipality and department within that municipality; however, since  department was missing in 95% of records in CHDES and 80% of records in ERTL, we excluded it from our linkage process. Thus we conduct linkage using given name, last name, municipality, and day, month, and year of death. We again use flat priors for the $\mathbf{m}$ and $\mathbf{u}$ parameters, but use 2000 Gibbs iterations with 200 discarded as burn-in to mirror the implementation in Sadinle 2017.

```{r}
include_graphics("../notes/figures/el_salvador/overlap_distribution_smallP.png")
```

```{r}
include_graphics("../notes/figures/el_salvador/m_posterior_smallP.png")
```

To mirror the original implementation, we constructed the comparison vectors using 4 levels of agreement for each field, according to the thresholds provided in Figure XX. This took ____ seconds for our proposed method and ____ for `BRL`, already demonstrating the increased efficiency of our method. However, we note that posterior distributions of several levels of the $\mathbf{m}$ and $\mathbf{u}$ parameters were nearly identical, and that of the $4^5 \times 2 = 2048$ possible agreement patterns, only ___ are realized in the data. This leads us to believe that such high number of aggreement levels creates unnecesary distictions in the data and makes the comparison vectors less interpretable. Therefore we re-ran our analysis with fewer agreement levels for each field (see Figure XX), and obtained analogous results. With ____ possible agreement patterns, ___ were realized in the data, and our proposed method became much faster, finishing in ___ seconds. Meanwhile `BRL` took ___ seconds, relatively unchanged from the first implementation. This demonstrates the way that the computational complexity of our method depends on the number of unique agreement patterns, and how significant computional gains can be made by simplifying the construction of the comparison vectors. We also note that estimates of the $\mathbf{m}$ and $\mathbf{u}$ paramters under each method are very similar, as shown in Figure XX.

## Violations of One-to-One Matching

Figure XX shows the posterior distribution for $D$, the number of duplicates found across file for both `fabl` and `BRL`. We see that `fabl` consistently overmatches within each Gibbs iteration when compared to `BRL`, which is to be expected because `BRL` explicitly prevents matches that violate one-to-one matching throughout the entire sampler. Most of such matchings are due to the randomness in the sampling procedure, and they occur sporadically throughout the sampler in such a way that does not measurably influence the eventual Bayes estimate.

Overall, both methods presented similar results. `fabl` yielded an initial Bayes estimate of ____ matches found across files, and after resolving matches that violated one-to-one requirements, yeilded ___ matches across files. This is acceptably close to the ___ matches found by `BRL`. The main reason for this discrepancy is the difference in how each model handles situations in which one file in $B$ has multiple plausible matches in $A$.

```{r}
include_graphics("../notes/figures/el_salvador/rosa_records.png")
```

```{r}
include_graphics("../notes/figures/el_salvador/bad_mixing.png")
```

The records shown in Figure XX provide one such instance. Note these records present a near violation of our assumption that there are no duplications within files; we continue to assume that records ___ and ____ in ERTL correspond to different individuals, but their records are nearly identical. Additionally, using the modified Levenstein distance from Sadinle 2017, the comparison vectors $\gamma_{2776, 825}$ and $\gamma_{2776, 826}$ are exactly identical.

Figure XX shows the values that $Z_{825}$ and $Z_{826}$ take on throughout the Gibbs sampler, and demonstrate how each method handles this situation. Under `parlr`, both records in $B$ match to the same record in $A$ throughout the Gibbs process, creating consistent violations of one-to-one matching. Under `BRL`, the Gibbs process creates one matching configuration stays there for a while. However, if one pair "unmatches," then the other record has a chance to latch on. Then,  Gibbs process is stuck with that matching status for a while, resulting in a Gibbs process with poor mixing. Additionally, `parlr` allows the modeller to inspect records with mulitple plausible matches, and if they desire, to then choose the record pairing with the highest posterior probability. `BRL` in contrast, in strictly enforcing one-to-one matching throughout the sampler, can lead to situations where none of the plausible matches reach the threshold to be identified through the Bayes estimate.

@@

We have presented a method for Bayesian record linkage that is feasible for large datasets. In particular, our hashing procedure and model assumptions allow for a linkage procedure whose computational complexity does not scale with the size of the larger dataset, making this method particularly powerful in linking records when one datafile is substantially smaller than the other.

In our case study, we included an exploration of how to conduct record linkage when modelling assumptions are not met in practice. We explored "one-to-many" scenarios in which
